Transcriptional dynamics of influenza virus infection in single cells

### 2017-09-01 paper/paper.tex
finished revision to Methods

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
\usepackage{natbib}

@@

\subsection{Computational analysis of single-cell mRNA sequencing data}
Sequencing data were aligned and a gene-cell-barcode matrix determined using the 10x Genomics software package, CellRanger v. 2.0.0. 
Wild-type and synonymous viral transcripts were identified using a custom python script, and these annotations appended to the gene-cell-barcode matrix for further analysis.
These matrices were then evaluated using the R package Monocle, version 2.4.0~\citep{Qiu:2017dw,Trapnell:2014kg}.
Cell barcodes associated with 2.5-fold fewer or more UMI counts than the sample median were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
In order to empirically determine a cutoff for potential lysis-derived influenza reads we used the 10 hour time point, which had the highest influenza transcript count.
The relative purity of barcodes was calculated out to the 35th percentile of influenza-associated reads.
These values were plotted against the number of UMIs attributed to influenza, and a curve was fitted using the loess function in R. 
The derivative of the resultant curve was calculated, and the threshold set at the minimum value wherein the derivative is less than 0 -- indicating the point below which there is a sacrifice in sample purity, indicating an increase in the relative proportion of extracellular-derived influenza reads.
Cells meeting this threshold were defined as infected, and those failing to do so were excluded from further analysis.
For each time point an additional number of cells, either 50 or the number of infected cells determined by our threshold above, whichever was larger, with no detectable influenza transcripts were also included for further analysis.

Using Monocle'?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influenza. 
Influenza infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
Candidate marker genes for t-distributed stochastic neighbor embedding (tSNE) dimensionality reduction that would allow us to separate these populations were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.

Analyses of host expression required alteration to the standard Monocle pipeline and subsetting of our dataset.
First, we only considered host genes in our analyses, as the massive contribution of viral variation could swamp out more nuanced effects.
Second, as influenza has the capacity to suppress the host transcriptome ~\citep{BercovichKinori:2016iw}, and moreover contributes significantly to the total number UMIs that are attributed to a cell, we decided to assign the cell's size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone. 
We excluded uninfected cells from these analyses, as well as those influenza-infected cells that lacked a gene from any of the eight segments. 
Lastly, cells which contained a fraction of influenza-derived reads of greater than 25\% were assessed at a value of 25\% such that they did not contribute an undue weight to our analysis. 
Genes whose expression co-varied with the fraction of mRNAs in a cell derived from influenza were then identified, using the Monocle function differentialGeneTest, after removing variance explained by sample to sample variation. 
Retaining those meeting a q-value threshold of 0.1, the expression of these genes was visualized on a per-cell basis, arranging cells in order of increasing percentage of reads derived from influenza.

As an additional analysis, we ranked cells within each sample according to the relative percentage of reads derived from influenza.
As above, we only considered cells expressing genes from all eight viral genomic segments, and determined size factors based on the host transcriptome alone.
Genes whose expression correlated with influenza ranking were then determined using the Monocle function differentialGeneTest, again removing variance explained by sample to sample variation and using a q-value threshold of 0.1. 
Cells were then arranged my ranking within each sample, in order of increasing percentage of reads derived from influenza within a sample, and differentially-regulated genes were visualized.

@@

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
\usepackage{natbib}
\usepackage{float}

\newfloat{suppfile}{thp}{lofsupfile}
%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
\floatname{suppfile}{Supplementary file}

@@

\subsection{Computational analysis of single-cell mRNA sequencing data}
Jupyter notebooks that perform all of the computational analyses are available at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript. If you are a reviewer who needs access, please contact the editor.}

Briefly, the raw deep sequencing data were processed using the 10X Genomics software package CellRanger (version 2.0.0).
The reads were aligned to a concatenation of the human and influenza virus transcriptomes.
The human transcriptome was generated by filtering genome assembly GRCh38 for protein coding genes defined in the GTF file GRCh38.87.
The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.
We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/align_and_annotate.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript. If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:alignandannotate}.

The annotated cell-gene matrix was then further analyzed in R, primarily using the Monocle package (version 2.4.0)~\citep{Qiu:2017dw,Trapnell:2014kg}.
The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/monocle_analysis.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript. If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:monocle}.

Briefly, this analysis was performed as follows.
For each sample, cell barcodes with 2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were considered to be potential doublets or inefficient/empty reactions, and were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C).
We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
The 10-hour sample has the most viral mRNA, and is expected to have the most lysis-associated influenza reads.
We therefore determined the threshold using just this sample (Figure~\ref{fig:viralbarcodes}C), and then applied this threshold to all samples.
This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
For subsequent analyses, we retained all infected cells and a subsample of uninfected cells (the greater of 50 or the number of infected cells for that sample).
The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the t-SNE clustering and other downstream analyses.

For the semi-supervised t-SNE clustering, we used Monocle's cell hierarchy function to bin cells into those with no viral mRNA, <2\% viral mRNA, between 2\% and 20\% viral mRNA, and >20\%. 
Candidate marker genes t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.

For the analyses of cellular genes that differed in expression as a function of the amount of viral mRNA, we only considered cells that expressed all 8 viral mRNAs to avoid effects driven simply by viral gene absence.
We also only considered cellular genes in the differential gene analysis, since viral gene expression will tautologically co-vary with the amount of viral mRNA.
Additionally, because influenza virus has the capacity to degrade or prevent the synthesis of host mRNAs~\citep{BercovichKinori:2016iw} and contributes significantly to the total number UMIs in some cells, we calculate size factors (a scalar value representing efficiency of UMI capture) based on cellular transcripts alone. 
Finally, we assigned all cells a ceiling fraction of mRNA from virus of 25\% so that a few extremely high-expressing cells did not dominate. 
Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation. 
\label{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.

@@

\clearpage

\begin{suppfile}
\caption{\label{suppfile:cellgenematrix}
The annotated cell-gene matrix in Matrix Market Format.}
\end{suppfile}

\begin{suppfile}
\caption{\label{suppfile:alignandannotate}
Jupyter notebook that runs CellRanger to align and annotate the reads.
The ZIP file also includes associated custom scripts run by this notebook.}
\end{suppfile}

\begin{suppfile}
\caption{\label{suppfile:monocle}
Jupyter notebook that analyzes the cell-gene matrix, primarily using Monocle.}
\end{suppfile}
